Synthesis of Linear Functional Programs

### 2021-05-09 article/article.tex
touch up

The elimination form for the exponential, written $\llet{ !u
= M}{ N}$, warrants the use of resource $A$ an unbounded
number of times in $N$ via the variable $u$:  
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
\]
Again, through the Curry-Howard correspondence, we can view the process
of finding a proof of a proposition in this linear logic as the process of synthesizing a
linear functional program of a given type.

\mypara{Resource-Management}
Coming from linear logic, we now have a new problem to solve:
When searching for a proof, how do we break the
contexts in the necessary way to allocate the correct resources
to prove each judgment the judgment we're proving depends on?
This is where the idea of \emph{input/output} contexts come in.
To solve this issue we will adopt the resource-management
approach
of~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:conf/lics/LiangM09}.
Instead of requiring non-deterministic guesses of resource splits
during proof search, we will make use of an \emph{input} and an
\emph{output} context, leading to the following general strategy: to
prove $A\tensor B$ using resources $\Delta$, prove $A$ with $\Delta$,
consuming some subset of $\Delta$, and produce as output the leftover
resources $\Delta'$; prove $B$ using $\Delta'$ and then output the
remaining resources $\Delta''$; finally, after having proven
$A\tensor B$, output $\Delta''$, for subsequent derivations.

\mypara{Sequent Calculus}
Inference rules in natural deduction (the ones we have considered so far)
are complicated to use in a proof-search strategy.
A more suited candidate is the equivalent \emph{sequent calculus} system
in which \emph{all} inference rules are read naturally in a \emph{bottom-up} manner.
The introduction and elimination rules from natural deduction disappear,
and its place is taken by right and left rules, respectively.
Right rules and introduction rules are almost identical, since introduction rules were
already read bottom-up. The inference rule for implication introduction ($\supset$ is used instead of $\rightarrow$), seen above in natural deduction under the
\emph{propositions as types} subsection, is the following right rule in sequent calculus:
\[
    \infer*[right=($\supset R$)]
    {\Gamma, u: A \vdash M : B}
    {\Gamma \vdash \lambda u. M : A \supset B}
\]
Left rules try to capture the same idea as elimination rules in natural deduction, but are altered
to reflect it bottom-up, instead of top-down (usual ``orientation'' of elimination rules).
The inference rule for the also seen above implication elimination, is the following left rule here:
\[
    \infer*[right=($\supset L$)]
    {\Gamma , u: A \supset B \vdash M : A \supset B \and \Gamma \vdash N : A}
    {\Gamma \vdash M\,N : B}
\]
We'll define our inference rules for synthesis under the sequent calculus system,
for it simplifies the work to be done in the synthetizer.

\mypara{Focusing}
Even with everything mentioned so far, non-determinism still pertains proof-search in logic, e,g., at any given point, many proof rules are applicable in general.
Focusing has been previously studied as a way to conduct proof search in linear logic~\cite{}
--
a method created to trim down the search space of valid proofs in
linear logic, by eagerly applying invertible rules (i.e.~rules whose
conclusion implies the premises), and then by ``focusing'' on a single
connective when no more direct (invertible) rules can be applied. If
the search is not successful, the procedure backtracks and another
connective is chosen as the focus.
%
Focusing eliminates all of the
``don't care'' non-determinism from proof search, since the order in
which invertible rules are applied does not affect the outcome of the
search, leaving only the non-determinism that pertains to unknowns (or
``don't know'' non-determinism), identifying precisely the points at
which backtracking search is necessary.

@@

With the use of input/output contexts, and the sequent calculus
formulation of the inference rules, we draw closer to an effective
strategy for bottom-up proof search and so of program synthesis.
%

However, the search process still encompasses significant
non-determinism. To address this we draw on the technique of
focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} as seen before.

@@

\mypara{Work Plan}
The end goal is to have a synthetizer capable of taking a type (possibly refined) as input,
and to produce (possibly interactively) a function or more (possibly recursive) with that same type.
We've already created a parser and a type-checker -- more than good preparation for the
development of the synthetizer, they will be useful to \emph{validate} the functions synthetized, and to implement the
synthesis of partial programs. The implementations are in Haskell, a functional language, since it's easier to reason about inference rules that way.

\mypara{Evaluation}
Validation and evaluation of our work is fairly simple. Simple questions such as \emph{does it typecheck}? \emph{can you synth}? \emph{how fast}?
are the core of our evaluation. In more concrete terms, we can compare our times to state of the art synthetizers, present a list of working examples
of programs synthesed by our synth, and according benchmarks. On the eventual addition of recursive functions, refinement types, and interactive synthesis,
we'll consider, respectively, other examples to showcase recursive functions synthesis, updated times for benchmarks with refinement types, and a display of
the possible user interaction in the user process, and the resulting function following it.

>>>>

The elimination form for the exponential, written $\llet{ !u
= M}{ N}$, warrants the use of resource $A$ an unbounded
number of times in $N$ via the variable $u$:  
\[
    \infer*[right=($\bang E$)]
    {\Gamma ; \Delta_1 \vdash M : \bang A \and \Gamma, u{:}A ; \Delta_2 \vdash N : C}
    {\Gamma ; \Delta_1, \Delta_2 \vdash \llet{ !u = M}{ N} : C}
\]
Again, through the Curry-Howard correspondence, we can view the process
of finding a proof of a proposition in linear logic as the process of synthesizing a
linear functional program of the given type.

\mypara{Resource-Management}
Be it from the perspective of proof \emph{checking} (i.e.~type checking) or
proof \emph{search} linear logic poses a key challenge when compared
to the non-linear setting:
When constructing a derivation (bottom-up), we are seemingly forced to
guess how to correctly split the linear context such that the
sub-derivations have access to the correct resources (e.g.~the
$\tensor I$ rule above).
%
%

To solve this issue we will adopt the resource-management
approach
of~\cite{DBLP:journals/tcs/CervesatoHP00,DBLP:conf/lics/LiangM09},
which generalizes the judgment  from $\Delta \vdash M : A$ to
$\Delta_I \ \Delta_O \vdash M : A$, where $\Delta_I$ is an input
context and $\Delta_O$ is an output context.
Instead of requiring non-deterministic guesses of resource splits
during proof search, we track which resources are used and which are
remaining via the two contexts, leading to the following general strategy: to
prove $A\tensor B$ using (input) resources $\Delta$, prove $A$ with
input context $\Delta$,
consuming some subset of $\Delta$, and produce as output the leftover
resources $\Delta'$; prove $B$ using $\Delta'$ as its input context and then output the
remaining resources $\Delta''$; finally, after having proven
$A\tensor B$, output $\Delta''$, for subsequent derivations.

\mypara{Sequent Calculus} Inference rules in natural deduction (the
ones we have considered so far) are ill-suited for bottom-up
proof-search since elimination rules work top-down and introduction
rules work bottom-up.  A more suited candidate is the equivalent
\emph{sequent calculus} system in which \emph{all} inference rules can
be understood naturally in a \emph{bottom-up} manner.  The
introduction and elimination rules from natural deduction disappear,
and their place is taken by right and left rules, respectively.
%
Right rules correspond exactly to the introduction rules of natural
deduction, which were already understood bottom-up. The inference rule
for implication introduction ($\supset$ is used instead of
$\rightarrow$), seen above in natural deduction under the
\emph{propositions as types} subsection, corresponds to the following
right rule in sequent calculus:
\[
    \infer*[right=($\supset R$)]
    {\Gamma, u: A \vdash M : B}
    {\Gamma \vdash \lambda u. M : A \supset B}
  \]

Intuitively, left rules act as the elimination rules of natural deduction, but are altered
to work bottom-up, instead of top-down.
The inference rule for the also seen above implication elimination, is
as follows:
\[
    \infer*[right=($\supset L$)]
    {\Gamma , u:B\vdash M : C \and \Gamma, u{:}A\supset B \vdash N : A}
    {\Gamma, u{:}A\supset B\vdash M\{(u\,N)/u\} : C}
  \]
  As implied by their name, left rules define how to decompose or make
  use of a connective on the left of the turnstile $\vdash$. To use an
  assumption of $A\supset B$ while attempting to show some proposition
  $C$ we produce a proof of $A$, which allows us to use an assumption
  of $B$ to prove $C$. In terms of the corresponding $\lambda$-terms,
  the $\supset L$ rule corresponds to applying the variable $u$ to the
  argument $N$ but potentially deep in the structure of $M$.

  We'll define our inference rules for synthesis under the sequent
  calculus system, for it simplifies the work to be done in the
  synthesizer.

  \mypara{Focusing}

  Even with everything mentioned so far, non-determinism is still
  present proof-search in logic, e,g., at any given point, many proof
  rules are applicable in general. The technique of focusing~\cite{}
  has been previously studied as a way to discipline proof search in
  linear logic-- a method created to trim down the search space of
  valid proofs in linear logic, by eagerly applying invertible rules
  (i.e.~rules whose conclusion implies the premises), and then by
  ``focusing'' on a single connective when no more direct (invertible)
  rules can be applied, that is, only applying rules that breakdown
  the connective under focus or its subformulas. If the search is not
  successful, the procedure backtracks and another connective is
  chosen as the focus.

  Focusing eliminates all of the ``don't care'' non-determinism from
  proof search, since the order in which invertible rules are applied
  does not affect the outcome of the search, leaving only the
  non-determinism that pertains to unknowns (or ``don't know''
  non-determinism), identifying precisely the points at which
  backtracking search is necessary.

@@

With the use of input/output contexts, and the sequent calculus
formulation of the inference rules, we draw closer to an effective
strategy for bottom-up proof search and so of program synthesis.
%
%
However, the search process still encompasses significant
non-determinism. To address this we draw on the technique of
focusing~\cite{10.1093/logcom/2.3.297,DBLP:conf/cade/ChaudhuriP05} as
discussed above.

@@

\subsection{Work Plan}

The end goal is to have a synthetizer capable of
taking a type (possibly refined) as input, and to produce (possibly
interactively) one or more functions with that type, if any exist.

As mentioned above, we will begin with (both partial and full)
synthesis for the linear $\lambda$-calculus. We will then extend the
language with additional type constructs (e.g. base types, records,
algebraic data types) and consider synthesis in that setting. Finally,
we will consider synthesis of recursive functions, and,
time-permitting, we will consider simple forms of type refinements
(e.g. arithmetic refinements).

We've already implemented a parser and a type-checker for a linear
$\lambda$-calculus using the resource management technique described
in previous section -- more than
good preparation for the development of the synthesizer, they will be
useful to \emph{validate} the synthesized functions, and to enable
the synthesis of partial programs.
%
The implementations have been carried out in Haskell since it's easier
to model and implement inference rules in a functional setting.

\mypara{Evaluation} Validation and evaluation of our work is fairly
simple. Simple questions such as \emph{does it typecheck}? \emph{can
  we synthesize}? \emph{how fast}?  are the core of our evaluation. In
more concrete terms, we can compare our times to state of the art
synthetizers, present a list of working examples of programs synthesized
by our framework, and corresponding benchmarks.

On the eventual addition of recursive functions, refinement types, and
interactive synthesis, we'll consider, respectively, other examples to
showcase recursive functions synthesis, updated times for benchmarks
with refinement types, and a display of the possible user interaction
in the user process, and the resulting function following it.
